---
title: "Front-leg sensorimotor loops"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Front-leg sensorimotor loops}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

A central claim of the BANC paper ([Bates et al. 2025](https://doi.org/10.1101/2025.07.31.667571))
is that motor control in *Drosophila* is **distributed** — many
behaviours are not orchestrated by a single brain command followed by
passive execution in the ventral nerve cord (VNC), but instead emerge
from local sensorimotor loops embedded inside individual VNC
neuromeres, coordinated by ascending and descending pathways.

The front-leg neuromere (T1) is a convenient place to see one of those
loops by eye: the leg's mechanosensory afferents enter through the
front-leg nerve, the motor neurons that drive the leg muscles exit
through the same nerve, and a population of small VNC intrinsic
neurons (the "single-leg-neuromere" cell class) sits between them. In
this article we use `bancr` to identify those intrinsic neurons by the
purely connectomic criterion of *being monosynaptic between left
front-leg sensory neurons and left front-leg motor neurons*, then
build a public Neuroglancer scene that shows the three groups
colour-coded.

Everything below runs against public BANC resources: annotations come
from the public GCS snapshot (no authentication required), and
synaptic connectivity comes from the read-only public CAVE datastack
`brain_and_nerve_cord_public`.

## Setup

Even though every step below runs against *public* BANC resources,
CAVE still requires a personal token tied to your Google account
before any of its endpoints will answer. If you haven't set one up
yet, run `banc_set_token()` once — it opens a browser, signs you in
via Google, and stashes the token for future sessions. The
[`bancr` README](https://natverse.github.io/bancr/#installation)
walks through the full setup. No production-access grant is needed
to follow this article; the token alone is enough.

```{r setup}
library(bancr)
library(dplyr)

# Pin every CAVE call in this article to the public read-only
# datastack so the analysis works without production access.
options(bancr.datastack = "brain_and_nerve_cord_public")
```

## Load the codex annotation table

`banc_codex_annotations()` returns a per-neuron pivot of the BANC
codex annotation table — the same one displayed at
<https://codex.flywire.ai/?dataset=banc>. By default it reads from a
cached public GCS snapshot, so this first call needs no CAVE token.

```{r codex}
codex <- banc_codex_annotations()
```

## Select left-front-leg sensory and motor pools

The codex carries the BANC annotation hierarchy (`super_class`,
`cell_class`, `cell_sub_class`, `cell_type`) plus laterality
(`side`), the sensed body part (`body_part_sensory`) and the motor
target (`body_part_effector`). We filter to neurons whose receptive or
motor field is the left front leg.

```{r pools}
sensory <- codex %>%
  filter(
    super_class == "sensory",
    side == "left",
    grepl("front_leg", body_part_sensory)
  )

motor <- codex %>%
  filter(
    grepl("motor", super_class),
    side == "left",
    grepl("front_leg", body_part_effector)
  )

c(sensory = nrow(sensory), motor = nrow(motor))
```

## Find intrinsic neurons that close the loop

To be on the loop, an intrinsic neuron must receive direct input from
at least one of the sensory neurons *and* send direct output to at
least one of the motor neurons. We use `banc_partner_summary()` to
pull the monosynaptic partner lists, one row per (query neuron →
partner) edge.

`version = 888` pins the call to the most recent materialisation
exposed by the public datastack. We deliberately pass `threshold = 0`
so the partners step keeps every candidate edge; the cell-class and
side filters below do the tightening.

```{r partners}
sensory_out <- banc_partner_summary(
  sensory$pt_root_id,
  partners = "outputs",
  threshold = 0,
  version = 888
)

motor_in <- banc_partner_summary(
  motor$pt_root_id,
  partners = "inputs",
  threshold = 0,
  version = 888
)

candidates <- intersect(
  as.character(sensory_out$post_id),
  as.character(motor_in$pre_id)
)
length(candidates)
```

The intersection of "downstream of any sensory neuron" and "upstream
of any motor neuron" is still loose — it includes brain interneurons,
contralateral pathways and projection neurons that just happen to
touch both ends. We tighten to a clean local-circuit population by
restricting to the `single_leg_neuromere` cell class on the same
(left) side.

```{r intrinsic}
intrinsic <- codex %>%
  filter(
    as.character(pt_root_id) %in% candidates,
    cell_class == "single_leg_neuromere",
    side == "left"
  )

nrow(intrinsic)
```

A quick breakdown by `cell_sub_class` gives a sense of the
sub-populations involved:

```{r breakdown}
intrinsic %>%
  count(cell_sub_class, sort = TRUE) %>%
  head(10)
```

## Render a sample of the loop in BANC space

Before we build the full interactive scene, here's a quick static
look at the geometry. We sample a handful of neurons per group, pull
their L2 skeletons (read against the same public datastack), and use
`bancr::banc_neuron_comparison_plot()` to render them against the VNC
neuropil. The plot's three colour ramps are blue (sensory) / red
(motor) / green (intrinsic), so they line up with the Neuroglancer
scene below.

```{r mini-render, fig.height = 6, fig.width = 8, fig.alt = "Sample left-front-leg sensory, intrinsic, and motor neurons in the VNC"}
set.seed(1)
sample_ids <- list(
  sensory   = sample(sensory$pt_root_id,   4),
  motor     = sample(motor$pt_root_id,     4),
  intrinsic = sample(intrinsic$pt_root_id, 4)
)

sens_skels <- banc_read_l2skel(sample_ids$sensory)
mot_skels  <- banc_read_l2skel(sample_ids$motor)
intr_skels <- banc_read_l2skel(sample_ids$intrinsic)

banc_neuron_comparison_plot(
  neuron1 = sens_skels, neuron1.info = "sensory (left front leg)",
  neuron2 = mot_skels,  neuron2.info = "motor (left front leg)",
  neuron3 = intr_skels, neuron3.info = "intrinsic (single_leg_neuromere, left)",
  region  = "vnc",
  alpha   = c(0.7, 0.7, 0.7)
)
```

The motor and intrinsic arbours hug the left T1 neuromere, and the
sensory afferents enter through the front-leg nerve and ramify in the
same volume — exactly the local-loop geometry the BANC paper
emphasises.

## Render the loop as a public Neuroglancer scene

`banc_scene()` returns the canonical public BANC Neuroglancer state.
We decode it, set the segmentation-proofreading layer's segment list
to the union of our three pools, colour each id by its `super_class`,
and shorten the resulting URL via `banc_shorturl()`.

```{r scene}
# Group → colour map (colour-blind-safe Okabe-Ito palette)
group_colours <- c(
  sensory   = "#0072B2",  # blue
  intrinsic = "#009E73",  # green
  motor     = "#D55E00"   # vermillion
)

# One row per neuron with its id and group colour
colourdf <- bind_rows(
  data.frame(ids = as.character(sensory$pt_root_id),
             col = group_colours[["sensory"]]),
  data.frame(ids = as.character(intrinsic$pt_root_id),
             col = group_colours[["intrinsic"]]),
  data.frame(ids = as.character(motor$pt_root_id),
             col = group_colours[["motor"]])
)

# Build the scene: load the base public BANC state, install the
# segment list on the proofreading layer, then layer the colour map
# on top. The base scene archives the proofreading layer and selects
# "BANC static" by default — flip both so the reader opens the link
# straight onto our coloured neurons.
sc <- fafbseg::ngl_decode_scene(banc_scene())
layer_idx <- match("segmentation proofreading",
                   sapply(sc$layers, `[[`, "name"))
sc$layers[[layer_idx]]$segments <- colourdf$ids
sc$layers[[layer_idx]]$archived <- FALSE
sc$selectedLayer <- list(visible = TRUE,
                         layer = "segmentation proofreading")

scene_url <- fafbseg::ngl_add_colours(
  as.character(sc),
  colourdf,
  layer = "segmentation proofreading"
)

short_url <- banc_shorturl(scene_url)
short_url
```

```{r link, echo = FALSE, results = "asis"}
cat(sprintf("[Open the front-leg sensorimotor scene in Spelunker](%s)\n",
            short_url))
```

The Neuroglancer scene shows the **`r nrow(sensory)`** sensory neurons
in blue, the **`r nrow(intrinsic)`** single-leg-neuromere intrinsic
neurons in green, and the **`r nrow(motor)`** motor neurons in
vermillion. Switching on a neuropil overlay makes it obvious that the
whole loop sits inside the T1 left neuromere — exactly the local
sensorimotor module the BANC paper highlights.

## Exporting the loop for downstream analysis

A common follow-up is to share the full list of root IDs and their
annotations with collaborators (e.g. for proofreading, for matching
against another connectome, or for use in a non-R workflow). The
analysis output is a single tidy data frame — bind the three pools,
keep the annotation columns you care about, and write to CSV.

```{r export, eval = FALSE}
loop_selection <- dplyr::bind_rows(
  sensory   |> dplyr::mutate(group = "sensory"),
  intrinsic |> dplyr::mutate(group = "intrinsic"),
  motor     |> dplyr::mutate(group = "motor")
) |>
  dplyr::select(
    group, pt_root_id,
    super_class, cell_class, cell_sub_class, cell_type,
    side, body_part_sensory, body_part_effector,
    nerve, hemilineage
  )

write.csv(loop_selection,
          file = "~/Downloads/front_leg_sensorimotor_loop.csv",
          row.names = FALSE)
```

## What next

This is a single, manually-defined slice of the distributed-control
architecture characterised in Bates et al. 2025. The same recipe — a
codex filter for the two endpoints, a partners-API intersection, a
cell-class filter for the intrinsic pool — extends naturally to other
body parts (`body_part_sensory` / `body_part_effector`), other sides,
and other intrinsic cell classes (`sequential_leg_neuromeres`,
`intrasegmental`, …). For whole-CNS analyses, swap the per-query
`banc_partner_summary()` calls for the precomputed
`banc_edgelist()` / `banc_edgelist_split()` snapshots (also public,
also GCS-backed) and join against the codex table in `dplyr`.
